Practical Parametricity for GADTs

### 2020-09-27 20-gadts/inf-rules.tex
revised Sections 1 and 2

\new{Maybe develop our theory for {\em any} $\lambda \geq \omega_1$,
  and then specialize to $\omega_1$ when discussing GADTs? Can we do
  that? It seems we really use properties of $\oCPO$ to get that
  interpretations of $\Nat$-types are well-defined.}

@@

{\color{red} We should ask what, intuitively, it means to have
  parametricity for a calculus with GADTs. In particular it should
  mean that any polymorphic function that takes a GADT as input should
  operate uniformly over all type instances of the GADT. That means
  that the GADT must itself be ``parametric'', i.e., uniform in its
  type indices. But what does this mean? For ADTs and nested types it
  means that the datatype elements are constructed in exactly the same
  ways at each type instance. For GADTs, it means that as well, but
  here we also have to take into account which instances are actually
  inhabited, since not all type instances of a GADT are inhabited.
  [Of course we did not have to do this for ADTs and nested types
    because {\em all} instances of ADTs and nested types are
    inhabited.] So, intuitively, we consider a GADT to be
  ``parametric'' if 1) the corresponding instances of the GADT are
  inhabited for for every instantiation of its type parameters
  $\ol\alpha$, and 2) the objects for each instantiation of the
  $\alpha$s are constructed in exactly the same way. That is, which
  instances are inhabited, as well as how the data in each of those
  instances are constructed, is type independent. We capture this in
  our calculus by requiring that, for $\Gamma;\Phi \vdash
  (\Lan^{\ol{\alpha^0}}_{\ol K} F)\, \ol A$, we have that
  $\emptyset;\ol\beta \vdash K$, and that $K$ does not involve any
  constants. (The latter ensures that a GADT with a constructor of
  type $\alpha \to Int$ is not parametric, which makes sense since
  there will be more ways to construct Int instances of the GADT than
  ways to construct its other type instances.) Further, to ensure that
  the categorical Lans interpreting $\Lan$-constructs are
  $\omega$-cocontinuous we also require that $K$ is a polynomial
  functor in the $\alpha$s (that necessarily will not involve any
  constant functors). This means that our constructors have types of
  the form $F\ol\beta \to F(K\ol\beta)$ with no constants in $K$, so
  that exactly the corresponding instances of the GADT are inhabited
  for every instantiation of the $\alpha$s, and the objects for each
  instantiation of the $\alpha$s are constructed in exactly the same
  way. That is, the inhabited instances and the construction of the
  data in those instances are type independent.}

@@

For each $k \ge 0$, we assume countable sets $\tvars^k$ of \emph{type
  constructor variables of arity $k$} and $\fvars^k$ of
\emph{functorial variables of arity $k$}, all mutually disjoint.
%disjoint for distinct $k$ and disjoint from each other.
The sets of all type constructor variables and functorial variables
are $\tvars = \bigcup_{k \ge 0} \tvars^k$ and $\fvars = \bigcup_{k \ge
  0} \fvars^k$, respectively, and a \emph{type variable} is any
element of $\tvars \cup \fvars$.  We use lower case Greek letters for
type variables, writing $\phi^k$ to indicate that $\phi \in \tvars^k
\cup \fvars^k$, and omitting the arity indicator $k$ when convenient,
unimportant, or clear from context. We reserve letters from the
beginning of the alphabet to denote type variables of arity $0$, i.e.,
elements of $\tvars^0 \cup \fvars^0$. We write $\overline{\zeta}$ for
either a set $\{\zeta_1,...,\zeta_n\}$ of type constructor variables
or a set of functorial variables when the cardinality $n$ of the set
is unimportant or clear from context. If $\Pos\,$ is a set of type
variables we write $\Pos, \overline{\phi}$ for $\Pos\, \cup
\overline{\phi}$ when $\Pos\, \cap \overline{\phi} = \emptyset$.  We
omit the vector notation for a singleton set, thus writing $\phi$,
instead of $\overline{\phi}$, for $\{\phi\}$.
\begin{dfn}
Let $V$ be a finite subset of\, $\tvars$, $\Pos$ be a finite
subset of\, $\fvars$, $\overline{\alpha}$ be a finite subset of\,
$\fvars^0$ disjoint from $\Pos$, and $\phi^k \in \fvars^k
\setminus \Pos$.  The set $\mcF^\Pos(V)$ of {\em functorial
  expressions} over $\Pos$ and $V$ are given by
\begin{align*}
  \mcF^\Pos(V) \; ::= \; 
  & \hspace*{0.15in}
\zerot \mid \onet 
\mid \Nat^\Pos \, \mcF^\Pos (V) \;
\mcF^\Pos (V) 
\mid \Pos\; \ol{\mcF^\Pos(V)}  \,
\mid V\, \ol{\mcF^\Pos(V)}  
\mid \mcF^{\Pos}(V) + \mcF^\Pos(V)\\
&\mid \mcF^{\Pos}(V) \times \mcF^\Pos(V) \mid \left(\mu
\phi^{~k}. \lambda \alpha_1...\alpha_k.  
\mcF^{\Pos,\alpha_1,...,\alpha_k,\phi}(V)\right)
\ol{\mcF^{\Pos}(V)}\\
&\mid (\Lan^{\ol\alpha}_{\ol{\F^{\ol\alpha}}} \F^{\Pos,\ol\alpha})\ol{\F^\Pos}
\end{align*}
\end{dfn}
\noindent
A \emph{type} over $\Pos$ and $V$ is any element of $\F^\Pos(V)$. The
difference with~\cite{jgj20} here lies solely in the incorporation of
functorial expressions constructed from $\Lan$.

@@

In the rule for $L_{\ol{\alpha}}x.t$, the $L$ operator binds all
occurrences of the type variables in $\ol{\alpha}$ in the type of the
term variable $x$ and in the body $t$, as well as all occurrences of
$x$ in $t$. In the rule for $t_{\ol K} s$ there is one functorial
expression in $\ol K$ for every functorial variable in $\ol
\alpha$. In the rule for $\map^{\ol{F},\ol{G}}_H$ there is one
functorial expression $F$ and one functorial expression $G$ for each
functorial variable in $\ol\phi$. Moreover, for each $\phi^k$ in
$\ol\phi$ the number of functorial variables in $\ol\beta$ in the
judgments for its corresponding functorial expresssions $F$ and $G$ is
$k$. In the rules for $\tin_H$ and $\fold^F_H$, the functorial
variables in $\ol{\beta}$ are fresh with respect to $H$, and there is
one $\beta$ for every $\alpha$. (Recall from above that, in order for
the types of $\tin_H$ and $\fold^F_H$ to be well-formed, the length of
$\alpha$ must equal the arity of $\phi$.) In the rule for $\int_{\ol
  K,H}\,t$, there is one functorial expression $A$ for every
functorial variable in $\ol \alpha$, and in the rule for
$\partial^{G,\ol K}_F\,\eta\,t$, there is one functorial expression
$A$ for every functorial expression in $\ol K$ (and hence for every
functorial variable in $\ol \beta$).

@@

\new{Sum and product intro and elim rules should be annotated with
  constituent types for consistency?

We should have a computation rule along the lines of: If $\eta :
\Nat^{\ol\alpha} F \,G[\ol{\beta := K}]$ then
\[\begin{array}{ll}
& (\partial_F^{G,\ol K} \eta)_{\ol{K[\ol{\alpha := A}]}} \circ (\int_{K,F})_{\ol
  A}\; \rightarrow \; \eta_{\ol A}\\
: & F[\ol{\alpha := A}] \to G[\ol{\beta := K[\ol{\alpha :=A}]}]\\
= & F[\ol{\alpha := A}] \to G[\ol{\beta := K}][\ol{\alpha := A}]
\end{array}\]
This will appear as a computational property of the term
interpretations.}

@@

@@

{\color{red} NOT NEEDED ANYMORE??
To accommodate GADTs, we will need to transition Reynolds' approach
from a $\set$-based semantics to a semantics based on
$\omega$-complete partial orders. We denote the category of
$\omega$-complete partial orders ($\oCPO$s) and their {\color{blue}
  monotone?} sup-preserving morphisms by $\omega$CPO. The underlying
set of an $\oCPO$ $A$ is denoted $|A|$. The category $\oCPOR$ of
$\oCPO$ relations has as its objects triples $(A,B,R)$, where $A, B:
\oCPO$; $R: \rel(|A|, |B|)$; {\color{blue} $(a,b) \in R$, $a \leq a'$,
  and $b \leq b'$ imply $(a',b') \in R$; so next part is redundant?}
and $(\bigvee_{i < \omega} a_i, \bigvee_{i < \omega} b_i) \in R$
whenever $(a_i)_{i < \omega}$ and $(b_i)_{i < \omega}$ are chains in
$A$ and $B$, respectively, such that $(a_i,b_i) \in R$ for all
$i$. The category $\oCPOR$ has as its morphisms from $(A,B,R)$ to
$(A',B',R')$ pairs $(f : A \to A',g : B \to B')$ of morphisms in
$\oCPO$ such that $(f\,a, g\,b) \in R'$ whenever $(a,b) \in R$. We
note that if $(f,g) : (A,B,R) \to (A',B',R')$ and $(a_i)_{i < \omega}$
and $(b_i)_{i < \omega}$ are chains in $A$ and $B$, respectively, then
$ (f \, (\bigvee_{i < \omega} a_i), g \, (\bigvee_{i < \omega} b_i)) =
(\bigvee_{i < \omega} (f a_i), \bigvee_{i < \omega} (g b_i)) \in R'$
necessarily holds.}  We write $R : \oCPOR(A,B)$ in place of $(A,B,R):
\oCPOR$ when convenient. If $R : \oCPOR(A,B)$ then we write $\pi_1 R$
and $\pi_2 R$ for the {\em domain} $A$ of $R$ and the {\em codomain}
$B$ of $R$, respectively. If $A : \oCPO$, then we write $\Eq_A =
(A,A,\Eq_{|A|})$ for the {\em equality relation} on $A$.

To adapt Reynolds' approach, we first inductively define, for each
type, an object interpretation in $\oCPO$ and a relational
interpretation in $\oCPOR$.  Next, we show that these interpretations
satisfy both an Identity Extension Lemma (Theorem~\ref{thm:iel}) and
an Abstraction Theorem (Theorem~\ref{thm:gen-at}) appropriate to the
$\oCPO$ setting. The key to proving our Identity Extension Lemma is a
familiar ``cutting down'' of the interpretations of universally
quantified types to include only the ``parametric'' elements; as
in~\cite{jgj20}, the relevant types of the calculus defined above are
the (now richer) $\Nat$-types.  The requisite cutting down requires
that the object interpretations of our types in $\omega$CPO are
defined simultaneously with their relational interpretations in
$\oCPOR$.  We give the object interpretations for our types in
Section~\ref{sec:obj-interp} and give their relational interpretations
in Section~\ref{sec:rel-interp}.  While the former are relatively
straightforward, the latter are less so, mainly because of the
cocontinuity conditions, adapted from the $\set$-based setting
of~\cite{jgj20}, that must hold if they are to be well-defined. We
develop these conditions in Section~\ref{sec:rel-interp}, which
separates Definitions~\ref{def:set-sem} and~\ref{def:rel-sem} in
space, but otherwise has no impact on the fact that they are given by
mutual induction.

@@

The object interpretations of the types in our calculus will be
$\omega$-cocontinuous functors between categories of
$\omega$-cocontinuous functors on categories constructed from the
locally $\omega$-presentable category $\oCPO$.  We therefore begin
by recording some important facts about locally $\omega$-presentable
categories and functors on them, and verifying verify the properties
needed to interpret our syntax.

\subsubsection{Preliminaries}

\new{Perhaps have as preliminaries to entire paper.}

A category is \emph{small} if its collection of morphisms is a set.
It is \emph{locally small} if, for any two objects $A$ and $B$, the
collection of morphisms from $A$ to $B$ is a set. A {\em small
  (co)limit} in a category $\mcC$ is a (co)limit of a diagram $F :
\mcA \to \mcC$, where $\mcA$ is a small category. A category $\mcC$
is {\em (co)complete} if it has all small (co)limits.

A poset $\mcD = (D,\leq)$ is \emph{$\omega$-directed} if every
countable subset of $D$ has a supremum. When $\mcD$ is considered
as a category, we write $d \in \mcD$ to indicate that $d$ is an object
of $\mcD$ (i.e., $d \in D$). An {\em $\omega$-directed colimit} in a
category $\mcC$ is a colimit of a diagram $F : {\mathcal D} \to \mcC$,
where $\mathcal D$ is an $\omega$-directed poset. A category $\mcC$
is {\em $\omega$-cocomplete} if it has all $\omega$-directed
colimits; a {\em cocomplete} category is one that has all colimits.

If $\mcA$ and $\mcC$ are $\omega$-cocomplete, then a functor $F :
\mcA \to \mcC$ is {\em $\omega$-cocontinuous} if it preserves
$\omega$-directed colimits. We write $[\A,\C]_{\omega}$ for the
category of $\omega$-cocontinuous functors from $\A$ to $\C$, and
$\C^\A$ for the category of {\em all} functors from $\A$ to $\C$.
Since (co)limits of functors are computed pointwise, $C^\A$ has all
(co)limits that $\C$ has, and (co)limits of (co)continuous functors
are again (co)continuous. It follows that $[\A,\C]_{(\omega)}$ is
($\omega$-)(co)complete whenever $\C$ is. In particular, if $F :
[\C, \C]_{(\omega)}$ for a(n) ($\omega$-)cocomplete category $\C$,
then we can define $F^0 = \Id$, $F^{\alpha + 1} = F \circ F^\alpha$ at
successor ordinals, and $F^{\bigcup_{\beta < \alpha} \beta} =
\colim{\beta < \alpha} F^\beta$ at limit ordinals.

If $\mcA$ is locally small, then an object $A$ of $\mcA$ is {\em
  $\omega$-presentable} if the functor $\mathsf{Hom}_\mcA(A, -) :
\mcA \to \Set$ preserves $\omega$-directed colimits, i.e., if for
every $\omega$-directed poset $\mathcal D$ and every functor $F :
{\mathcal D} \to \A$, there is a canonical isomorphism $\colim{d \in
  \mcD}{\mathsf{Hom}_\mcA(A,Fd)} \simeq \mathsf{Hom}_{\mcA}(A,
\colim{d \in \mcD}{Fd})$. A locally small category $\mcA$ is {\em
  $\omega$-accessible} if it is $\omega$-cocomplete and has a set
$\mcA_0$ of $\omega$-presentable objects such that every object is
an $\omega$-directed colimit of objects in $\mcA_0$; a locally small
category is {\em locally $\omega$-presentable} if it is
$\omega$-accessible and cocomplete.

The category $\oCPO$ is locally presentable; its presentable objects
are precisely the finite sets.
%The category $\oCPO$ is locally $\omega$-presentable (but not
%locally finitely presentable); see Examples~1.18(2)
%of~\cite{ar94}). Its $\omega$-presentable objects are precisely the
%$\omega$CPOs cardinality less than $\omega$, i.e., the countable
%$\omega$CPOs.
In the next subsection we will interpret type variables in $\tvars^k
\cup \fvars^k$ as elements of $[\oCPO^k,\oCPO]_{\omega}$; the
following special cases of standard results (see, e.g.,~\cite{ar94})
will therefore be critical to deducing important properties of our
object interpretations of types:
\begin{prop}\label{prop:prelims}
\begin{enumerate}
\item If $\C_1,...,\C_n$ are locally $\omega$-presentable categories
  then so is $\C_1 \times ... \times \C_n$. Moreover, the presentable
  objects of $\C_1 \times ... \times \C_n$ are exactly the tuples of
  the form $(P_1,...,P_n)$, where, for each $i = 1,...,n$, the object
  $P_i$ is presentable in $\C_i$.
\item If $\mathcal{A}$ is $\omega$-accessible and $\mathcal{C}$ is
  $\lambda$-cocomplete, then the category $[\mathcal{A},
  \mathcal{C}]_{\omega}$ is naturally equivalent to the category
  $\mathcal{C}^{\mathcal{A}_0}$.
\item If $\mathcal{C}$ is locally $\omega$-presentable and
  $\mathcal{A}_0$ is essentially small, then
  $\mathcal{C}^{\mathcal{A}_0}$ is locally $\omega$-presentable.
\end{enumerate}
\end{prop}
\noindent
Together, the statements in Proposition~\ref{prop:prelims} give that
if $\mathcal{A}$ and $\mathcal{C}$ are locally $\omega$-presentable,
then $[\mathcal{A}, \mathcal{C}]_{\omega}$ is naturally equivalent
to $\mathcal{C}^{\mathcal{A}_0}$, and thus is
$\omega$-presentable. Thus, for all $k_1,...,k_n \in \nat^n$,
$[\mathcal{A}^{k_1}, \mathcal{C}]_{\omega} \times ... \times
[\mathcal{A}^{k_n}, \mathcal{C}]_{\omega}$ is locally
$\omega$-presentable, and therefore $[[\mathcal{A}^{k_1},
    \mathcal{C}]_{\omega} \times ... \times [\mathcal{A}^{k_n},
    \mathcal{C}]_{\omega}, \mathcal{C}]_{\omega}$ is as
well. Taking both $\A$ and $C$ to be $\oCPO$ --- as we will to ensure
that the fixpoints interpreting $\mu$-types in $\oCPO$ exist --- we
have
\begin{prop}\label{prop:nat-interps-sets}
  For all $k_1,...,k_n \in \nat^n$,
  \[[[\oCPO^{k_1}, \oCPO]_{\omega}
  \times ... \times [\oCPO^{k_n}, \oCPO]_{\omega},
  \oCPO]_{\omega}\] is locally $\omega$-presentable.
\end{prop}

\subsubsection{Object Interpretations}

To define the object interpretations of the types in
Definition~\ref{def:wftypes} we must first interpret their
variables. We have:

\begin{dfn}\label{def:set-env}
A {\em $\oCPO$ environment} maps each type variable in $\tvars^k \cup
\fvars^k$ to an element of $[\oCPO^k,\oCPO]_{\omega}$.  A morphism
$f : \rho \to \rho'$ for set environments $\rho$ and $\rho'$ with
$\rho|_\tvars = \rho'|_\tvars$ maps each type constructor variable
$\psi^k \in \tvars$ to the identity natural transformation on $\rho
\psi^k = \rho'\psi^k$ and each functorial variable $\phi^k \in \fvars$
to a natural transformation from the $k$-ary functor $\rho \phi^k$ on
$\oCPO$ to the $k$-ary functor $\rho' \phi^k$ on $\oCPO$.  Composition
of morphisms on set environments is given componentwise, with the
identity morphism mapping each set environment to itself. This gives a
category of set environments and morphisms between them, which we
denote $\oCPOenv$.
\end{dfn}

>>>>

{\color{blue} We should ask what, intuitively, it means to have
  parametricity for a calculus with GADTs. In particular it should
  mean that any polymorphic function that takes a GADT as input should
  operate uniformly over all type instances of the GADT. That means
  that the GADT must itself be ``parametric'', i.e., uniform in its
  type indices. But what does this mean? For ADTs and nested types it
  means that the datatype elements are constructed in exactly the same
  ways at each type instance. For GADTs, it means that as well. That
  is, we require related elements of a GADT to be constructed with the
  same sequence of constructors acting on related data. This entails
  that which instances are inhabited, as well as how the data in each
  of those instances are constructed, is type independent.}

@@

\new{Maybe develop our theory for {\em any} $\lambda \geq \omega$, and
  then specialize to $\omega_1$ (for $\omega$CPO) when discussing
  GADTs in future work? Can we do that? It seems we really use
  properties of $\oCPO$ to get that interpretations of $\Nat$-types
  are well-defined. It's not clear how to put $\omega$CPO structures
  on the interpretations of $\Lan$-types.}

@@

For each $k \ge 0$, we assume countable sets $\tvars^k$ of \emph{type
  constructor variables of arity $k$} and $\fvars^k$ of
\emph{functorial variables of arity $k$}, all mutually disjoint.
%disjoint for distinct $k$ and disjoint from each other.
The sets of all type constructor variables and functorial variables
are $\tvars = \bigcup_{k \ge 0} \tvars^k$ and $\fvars = \bigcup_{k \ge
  0} \fvars^k$, respectively, and a \emph{type variable} is any
element of $\tvars \cup \fvars$.  We use lower case Greek letters for
type variables, writing $\phi^k$ to indicate that $\phi \in \tvars^k
\cup \fvars^k$, and omitting the arity indicator $k$ when convenient,
unimportant, or clear from context. We reserve letters from the
beginning of the alphabet to denote type variables of arity $0$, i.e.,
elements of $\tvars^0 \cup \fvars^0$. We write $\overline{\zeta}$ for
either a set $\{\zeta_1,...,\zeta_n\}$ of type constructor variables
or a set of functorial variables when the cardinality $n$ of the set
is unimportant or clear from context. If $\Pos\,$ is a set of type
variables we write $\Pos, \overline{\phi}$ for $\Pos\, \cup
\overline{\phi}$ when $\Pos\, \cap \overline{\phi} = \emptyset$.  We
omit the vector notation for a singleton set, thus writing $\phi$,
instead of $\overline{\phi}$, for $\{\phi\}$.
\begin{dfn}
Let $V$ be a finite subset of\, $\tvars$, $\Pos$ be a finite
subset of\, $\fvars$, $\overline{\alpha}$ be a finite subset of\,
$\fvars^0$ disjoint from $\Pos$, and $\phi^k \in \fvars^k
\setminus \Pos$.  The set $\mcF^\Pos(V)$ of {\em functorial
  expressions} over $\Pos$ and $V$ are given by
\begin{align*}
  \mcF^\Pos(V) \; ::= \; 
  & \hspace*{0.15in}
\zerot \mid \onet 
\mid \Nat^\Pos \, \mcF^\Pos (V) \;
\mcF^\Pos (V) 
\mid \Pos\; \ol{\mcF^\Pos(V)}  \,
\mid V\, \ol{\mcF^\Pos(V)}  
\mid \mcF^{\Pos}(V) + \mcF^\Pos(V)\\
&\mid \mcF^{\Pos}(V) \times \mcF^\Pos(V) \mid \left(\mu
\phi^{~k}. \lambda \alpha_1...\alpha_k.  
\mcF^{\Pos,\alpha_1,...,\alpha_k,\phi}(V)\right)
\ol{\mcF^{\Pos}(V)}\\
&\mid (\Lan^{\ol\alpha}_{\ol{\F^{\ol\alpha}}} \F^{\Pos,\ol\alpha})\ol{\F^\Pos}
\end{align*}
\end{dfn}
\noindent
A \emph{type} over $\Pos$ and $V$ is any element of $\F^\Pos(V)$. The
difference with~\cite{jgj20} here lies solely in the incorporation of
functorial expressions constructed using $\Lan$.

@@

\new{Sum and product intro and elim rules should be annotated with
  constituent types for consistency?}

In the rule for $L_{\ol{\alpha}}x.t$, the $L$ operator binds all
occurrences of the type variables in $\ol{\alpha}$ in the type of the
term variable $x$ and in the body $t$, as well as all occurrences of
$x$ in $t$. In the rule for $t_{\ol K} s$ there is one functorial
expression in $\ol K$ for every functorial variable in $\ol
\alpha$. In the rule for $\map^{\ol{F},\ol{G}}_H$ there is one
functorial expression $F$ and one functorial expression $G$ for each
functorial variable in $\ol\phi$. Moreover, for each $\phi^k$ in
$\ol\phi$ the number of functorial variables in $\ol\beta$ in the
judgments for its corresponding functorial expresssions $F$ and $G$ is
$k$. In the rules for $\tin_H$ and $\fold^F_H$, the functorial
variables in $\ol{\beta}$ are fresh with respect to $H$, and there is
one $\beta$ for every $\alpha$. (Recall from above that, in order for
the types of $\tin_H$ and $\fold^F_H$ to be well-formed, the length of
$\alpha$ must equal the arity of $\phi$.)  {\color{red} In the rule
  for $\int_{\ol K,H}$, ANYTHING NEEDED HERE?}

@@

\new{We should have a computation rule along the lines of: If $\eta :
\Nat^{\ol\alpha} F \,G[\ol{\beta := K}]$ then
\[\begin{array}{ll}
& (\partial_F^{G,\ol K} \eta)_{\ol{K[\ol{\alpha := A}]}} \circ (\int_{K,F})_{\ol
  A}\; \rightarrow \; \eta_{\ol A}\\
: & F[\ol{\alpha := A}] \to G[\ol{\beta := K[\ol{\alpha :=A}]}]\\
= & F[\ol{\alpha := A}] \to G[\ol{\beta := K}][\ol{\alpha := A}]
\end{array}\]
This will appear as a computational property of the term
interpretations.}

@@

We denote the category of sets and functions by $\set$. The category
$\rel$ has as its objects triples $(A,B,R)$ where $R$ is a relation
between the objects $A$ and $B$ in $\set$, i.e., a subset of $A \times
B$, and has as its morphisms from $(A,B,R)$ to $(A',B',R')$ pairs $(f
: A \to A',g : B \to B')$ of morphisms in $\set$ such that $(f a,g\,b)
\in R'$ whenever $(a,b) \in R$. We write $R : \rel(A,B)$ in place of
$(A,B,R)$ when convenient.  If $R : \rel(A,B)$ we write $\pi_1 R$ and
$\pi_2 R$ for the {\em domain} $A$ of $R$ and the {\em codomain} $B$
of $R$, respectively.  If $A : \set$, then we write $\Eq_A =
(A,A,\{(x,x)~|~ x \in A\})$ for the {\em equality relation} on $A$.

@@

In the remainder of this section we first inductively define, for each
type, an object interpretation in $\set$ and a relational
interpretation in $\rel$. In Section~\ref{sec:iel} we show that these
interpretations satisfy both an Identity Extension Lemma
(Theorem~\ref{thm:iel}) and an Abstraction Theorem
(Theorem~\ref{thm:gen-at}) appropriate to the GADT setting. The key to
proving our Identity Extension Lemma is a familiar ``cutting down'' of
the interpretations of universally quantified types to include only
the ``parametric'' elements; the relevant types of the calculus
defined above are the $\Nat$-types (which are now richer than those
of~\cite{jgj20}) and the $\Lan$-types. The requisite cutting down
requires that the object interpretations of our types in $\set$ are
defined simultaneously with their relational interpretations in
$\rel$. We give the object interpretations for our types in
Section~\ref{sec:obj-interp} and give their relational interpretations
in Section~\ref{sec:rel-interp}.  While the former are relatively
straightforward, the latter are less so because of the delicacy of
interpreting the $\Lan$-types. (The object and relational
interpretations of $\Lan$-types are not defined in the same
``parallel'' way, as are the two interpretations of the other types.)
Another subtle point is guaranteeing the cocontinuity conditions,
adapted to our richer setting from~\cite{jgj20}, that must hold if the
type interpretations are to be well-defined. We develop these
conditions in Section~\ref{sec:rel-interp}, which separates
Definitions~\ref{def:set-sem} and~\ref{def:rel-sem} in space, but
otherwise has no impact on the fact that these definitions are given
by mutual induction.

@@

The object interpretations of the types in our calculus will be
$\omega$-cocontinuous functors between categories of
$\omega$-cocontinuous functors between categories constructed from the
locally finitely presentable category $\set$~\cite{ar94}; the
relational interpretations will be similar for the locally finitely
presentable category $\rel$. The fact that functor categories of
locally finitely presentable categories are again locally finitely
presentable ensures that the fixpoints interpreting $\mu$-types in
$\set$ and $\rel$ exist, and thus that both the set and relational
interpretations of all of the types in Definition~\ref{def:wftypes}
are well-defined~\cite{jp19}.

To define the interpretations of the types in
Definition~\ref{def:wftypes} we must first interpret their variables.
Writing $[\C,\D]$ for the category of $\omega$-cocontinuous functors
from $\C$ to $\D$ for the locally finitely presentable categories $\C$
and $\D$, for our $\set$ interpretations we have:

\begin{dfn}\label{def:set-env}
A {\em $\set$ environment} maps each type variable in $\tvars^k \cup
\fvars^k$ to an element of\, $[\oCPO^k,\oCPO]$. A morphism $f : \rho
\to \rho'$ for set environments $\rho$ and $\rho'$ with $\rho|_\tvars
= \rho'|_\tvars$ maps each type constructor variable $\psi^k \in
\tvars$ to the identity natural transformation on $\rho \psi^k =
\rho'\psi^k$ and each functorial variable $\phi^k \in \fvars$ to a
natural transformation from the $k$-ary functor $\rho \phi^k$ on
$\oCPO$ to the $k$-ary functor $\rho' \phi^k$ on $\oCPO$.  Composition
of morphisms on set environments is given componentwise, with the
identity morphism mapping each set environment to itself. This gives a
category of set environments and morphisms between them, which we
denote $\oCPOenv$.
\end{dfn}

When convenient we identify a functor in $[\oCPO^0, \oCPO]$ with its
value on $\ast$ and consider a $\oCPO$ environment to map a type
variable of arity $0$ to an $\omega$-cocontinuous functor from
$\oCPO^0$ to $\oCPO$, i.e., to an $\oCPO$. If $\Phi =
\{\phi_1^{k_1},...,\phi_n^{k_n}\}$ and $\ol K = \{K_1,...,K_n\}$ where
$K_i : [\oCPO^{k_i},\oCPO]_{\omega}$ for $i = 1,...,n$, then we write
$\rho[\Phi := \ol K]$ for the $\oCPO$ environment $\rho'$ such that
$\rho' \phi_i = K_i$ for $i = 1,...,n$ and $\rho' \phi = \rho \phi$ if
$\phi \not \in \Phi$.  If $\rho$ is an $\oCPO$ environment, we write
$\Eq_\rho$ for the $\oCPOR$ environment (see
Definition~\ref{def:reln-env}) such that $\Eq_\rho v = \Eq_{\rho v}$
for every type variable $v$. The categories $\oCPORT_k$ and relational
interpretations appearing in the third clause of
Definition~\ref{def:set-sem} are given in full in
Section~\ref{sec:rel-interp}.

\begin{dfn}\label{def:set-sem}
The {\em object interpretation} $\oCPOsem{\cdot} : \F \to [\oCPOenv,
  \oCPO]$ is defined by
\begin{align*}
  \oCPOsem{\Gamma;\Phi \vdash \zerot}\rho &= 0\\
  \oCPOsem{\Gamma;\Phi \vdash \onet}\rho &= 1\\
  \oCPOsem{\Gamma; \emptyset
    \vdash \Nat^\Phi
    \,F\,G}\rho &= \{\eta : \lambda \ol{K}. \,\oCPOsem{\Gamma;
    \Phi \vdash
    F}\rho[\Phi := \ol K] 
      \Rightarrow \lambda \ol{K}.\,\oCPOsem{\Gamma; 
        \Phi \vdash G}\rho[\Phi := \ol K] \\ 
      &\hspace{0.3in}|~\forall \ol{K = (K^1, K^2, K^*) : \oCPORT_k}.\\
      &\hspace{0.4in}(\eta_{\overline{K^1}}, \eta_{\overline{K^2}})
      : \oCPORsem{\Gamma; \Phi \vdash F}\Eq_{\rho}[\Phi := \ol K]
      \rightarrow \oCPORsem{\Gamma; \Phi \vdash
        G}\Eq_{\rho}[\Phi := \ol K] \} \\
  \oCPOsem{\Gamma;\Phi \vdash \phi\ol{F}}\rho &=
  (\rho\phi)\,\ol{\oCPOsem{\Gamma;\Phi \vdash
    F}\rho}\\
  \oCPOsem{\Gamma;\Phi \vdash F+G}\rho &=
  \oCPOsem{\Gamma;\Phi \vdash F}\rho +
  \oCPOsem{\Gamma;\Phi \vdash G}\rho\\
  \oCPOsem{\Gamma;\Phi \vdash F \times G}\rho &=
  \oCPOsem{\Gamma;\Phi \vdash F}\rho \times
  \oCPOsem{\Gamma;\Phi \vdash G}\rho\\ 
  \oCPOsem{\Gamma;\Phi,\ol\gamma \vdash (\mu \phi.\lambda
    \ol{\alpha}. H)\ol{G}}\rho &= (\mu
    T^\oCPO_{H,\rho})\ol{\oCPOsem{\Gamma;\Phi,\ol\gamma \vdash G}\rho}\\
    \text{where } T^\oCPO_{H,\rho}\,F & = \lambda
  \ol{A}. \oCPOsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash
    H}\rho[\phi :=  F][\ol{\alpha := A}]\\
  \text{and } T^\oCPO_{H,\rho}\,\eta &= \lambda
  \ol{A}. \oCPOsem{\Gamma;\ol\gamma,\phi, \ol{\alpha} \vdash
    H}\id_\rho[\phi := \eta][\ol{\alpha := \id_{A}}] \\
  \oCPOsem{\Gamma;\Phi \vdash (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\rho &=
    \{ t : (\textit{Lan}_{\ol{\oCPOsem{\Gamma; \ol\alpha \vdash
          K}\rho[\ol{\alpha := \_}]}} \oCPOsem{\Gamma; \Phi, \ol\alpha
      \vdash F}\rho[\ol{\alpha := \_}]) \ol{\oCPOsem{\Gamma; \Phi
        \vdash A}\rho}\\
    &\hspace{0.3in}|~(t,t) \in \oCPORsem{\Gamma;\Phi \vdash
      (\Lan^{\ol{\alpha}}_{\ol{K}} F)\ol{A}}\Eq_\rho\}\\
\end{align*}
\end{dfn}
\new{WHERE DOES THIS ALL GO? Must explain $\Lan$s, say why they are
  functors, say they can be computed using colimits in $\set$ and
  $\rel$, define injections, etc. If $f : B \to B'$ in $\set$, then
  the functorial action of $\mathit{Lan}_{\ol K}\,F$ on $f$ is the
  unique morphism $(\mathit{Lan}_{\ol K}\,F)f$ such that
  \begin{equation}\label{def:functorial-action-lan}
(\mathit{Lan}_{\ol K}\,F)f \circ \iota_{X,g} = \iota'_{X, f \circ
      g}~~~MISSING~SOME~BARS
\end{equation}
Here, $\iota_{X,g}$ is the injection of $FX$ into $(\mathit{Lan}_{\ol
  K}\,F)\,B = \colim{X : \set_0,\, g : KX \to B}{FX}$ and
$\iota'_{X,g'}$ is the injection of $FX$ into $(\mathit{Lan}_{\ol
  K}\,F)\,B' = \colim{X : \set_0,\, g' : KX \to B'}{FX}$.}
